---
title: IPL 2025 Contribution Score Analysis
date: "2026-01-19"
categories: [cricket, data science, r]
---

In a [previous blog post](https://bfern.github.io/posts/2023-05-17-introducing-contribution-score/), I have introduced the metric *contribution score*, which looks at the game situation, opposition player and venue and tells you how well a player has performed for each delivery of a match. This allows you to compare performances between batters and bowlers, and understand the impact of wickets compared with run prevention. I have calculated this contribution score for all deliveries in the 2025 IPL season, and in this post I will provide some basic analysis of these values.

```{r}
#| label: setup
#| include: false
library(readr)
library(dplyr)
library(yaml)
library(DBI)
library(RPostgres)

config <- read_yaml(file.path(path.expand("~"), ".config", "config.yaml"))
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "cricket",
  host = config$cricketdb$host,
  port = 5432,
  user = config$cricketdb$user,
  password = config$cricketdb$password
)

balls_df <- dbGetQuery(
  con,
  "
  select
    b.matchid,
    b.inningsid,
    b.overid,
    b.deliveryid,
    b.striker,
    b.bowler
  from
    staging.balls b
  left join
    staging.matches m
  on
    b.matchid = m.matchid
  where
    (m.comp = 'IPL') and
    (m.date >= '2025-03-22') and
    (m.date <= '2025-06-03')
  ;
"
)
dbDisconnect(con)

batting_contributions_df <- read_csv(file.path(
  "data",
  "ipl25_batting_contributions.csv"
)) |>
  left_join(balls_df, by = c("matchid", "inningsid", "overid", "deliveryid"))
bowling_contributions_df <- read_csv(file.path(
  "data",
  "ipl25_bowling_contributions.csv"
)) |>
  left_join(balls_df, by = c("matchid", "inningsid", "overid", "deliveryid"))
```

We will start off with an interactive summary of the contribution scores per batter over all game situations. Toggle the tables to order by a different column value (simply click on the column).

```{r}
batting_contributions_df |>
  group_by(striker) |>
  summarise(
    num_innings = length(unique(matchid)),
    num_balls = n(),
    total_contr = sum(contr),
    .groups = "drop"
  ) |>
  mutate(
    mean_contr_per_innings = total_contr / num_innings,
    mean_contr_per_ball = total_contr / num_balls
  ) |>
  arrange(desc(total_contr)) |>
  head(20)
```

Lets now break this down into different phases of the game: powerplay, middle overs (overs 7-15) and death (overs 16-20).

Lets now see how the same table for bowlers.

```{r}
bowling_contributions_df |>
  group_by(bowler) |>
  summarise(
    num_innings = length(unique(matchid)),
    num_balls = n(),
    total_contr = sum(contr),
    .groups = "drop"
  ) |>
  mutate(
    mean_contr_per_innings = total_contr / num_innings,
    mean_contr_per_ball = total_contr / num_balls
  ) |>
  arrange(total_contr) |>
  head(20)
```

And lets also break this down into powerplay, middle overs and death phases.